Physics-based protein structure prediction and design using the confinement method

### 2012-10-31 Paper/paper.tex
before merge

The calculation of free energy differences is of central importance in the simulation of biochemical
systems.  The computation of the free energy difference between pairs of macromolecular
conformations is currently a particularly difficult and computationally expensive task with existing
methods. In this work, an improved \Alberto{what is the improvement} version of the confinement approach is used to calculate absolute
free energies of bio molecular systems. The method provides two main advantages: it does not require a reaction coordinate or
transition path and it is fast to compute. Throught out the article we show its use in different
challenging modeling problems. In particular, we show its use in predicting the conformational
preference of chamaleon sequences (sequences with high sequence identity and different
folds). This sequence dependent conformational preference set the stage for the use of the method in
protein design.

@@

This method is particularly well suited to pick out the native/native like structure from misfolded or decoys. We have
tested this using different models from CASP (critical assessment of structure
prediction). CASP is a blind test world wide competition in which different groups apply their
methods to predict the 3D structure of proteins given their sequence. This is done with a strict 3
week limit on each target (3 day for servers) and each group is allowed to submit 5 possible
structures (ranked from better to worst). In our role as assessors during the CASP8 and CASP9 \cite{MacCallum2011}
competition we observed no correlation at all between the ordering of structures submited by the
groups and the real ranking compared to an experimental model \cite{Kryshtafovych2011}}. The consequences of this go beyond
those five structures; the deeper meaning is that groups producing ensembles of structures are
generating structures that are better than the ones they submit, but they do not know about it. In
fact, when querrying different groups after the results were known, most groups agree that this is
the case. Beyond the CASP problem, this reflects on the modelers ability to correctly rank order
models in many different environments, from structures for drug design leads to designing more
stable proteins or peptide mimetics such as peptoids. One of the main culprits of this lack of
accuracy is the fact that ranking is often done via a potential energy function, which in many cases
lacks an entropy component. Other initiatives including knowledge based potentials do use some sort
of free energy to rank order, but its accuracy is not enough. Our method provides a physics based
solution to this problem. We have tried two experiments centered on CASP. First we tried to rank
order some structures from the previous CASP9 experiment. Our second test was to participate as a
metapredictor group in CASP10. In both methods, our ranking is determined as a free energy measure,
and it is compared with the ranking given by a geometrical comparison between native and submitted
models called GDT\_TS (Global distance test score \cite{Zemla2003}). 
GDT\_TS represents the average percentage of residues that are in close proximity in two
structures optimally superimposed using four different distance cutoffs (1, 2, 4 and 8 Å).

>>>>

The calculation of free energy differences is of central importance in the simulation of biochemical
systems.  The computation of the free energy difference between pairs of macromolecular
conformations is currently a particularly difficult and computationally expensive task with existing
methods. In this work, an improved \Alberto{Arijit: what is the improvement??} version of the confinement approach is used to calculate absolute
free energies of bio molecular systems. The method provides two main advantages: it does not require a reaction coordinate or
transition path and it is fast to compute. Throught out the article we show its use in different
challenging modeling problems. In particular, we show its use in predicting the conformational
preference of chamaleon sequences (sequences with high sequence identity and different
folds). This sequence dependent conformational preference set the stage for the use of the method in
protein design.

@@

This method is particularly well suited to pick out the native/native like structure from misfolded or decoys. We have
tested this using different models from CASP (critical assessment of structure
prediction). CASP is a blind test world wide competition in which different groups apply their
methods to predict the 3D structure of proteins given their sequence. This is done with a strict 3
week limit on each target (3 day for servers) and each group is allowed to submit 5 possible
structures (ranked from better to worst). In our role as assessors during the CASP8 and CASP9 \cite{MacCallum2011}
competition we observed no correlation at all between the ordering of structures submited by the
groups and the real ranking compared to an experimental model \cite{Kryshtafovych2011}. The consequences of this go beyond
those five structures; the deeper meaning is that groups producing ensembles of structures are
generating structures that are better than the ones they submit, but they do not know about it. In
fact, when querrying different groups after the results were known, most groups agree that this is
the case. Beyond the CASP problem, this reflects on the modelers ability to correctly rank order
models in many different environments, from structures for drug design leads to designing more
stable proteins or peptide mimetics such as peptoids. One of the main culprits of this lack of
accuracy is the fact that ranking is often done via a potential energy function, which in many cases
lacks an entropy component. Other initiatives including knowledge based potentials do use some sort
of free energy to rank order, but its accuracy is not enough. Our method provides a physics based
solution to this problem. We have tried two experiments centered on CASP. First we tried to rank
order some structures from the previous CASP9 experiment. Our second test was to participate as a
metapredictor group in CASP10. In both methods, our ranking is determined as a free energy measure,
and it is compared with the ranking given by a geometrical comparison between native and submitted
models called GDT\_TS (Global distance test score \cite{Zemla2003}). 
GDT\_TS represents the average percentage of residues that are in close proximity in two
structures optimally superimposed using four different distance cutoffs (1, 2, 4 and 8 Å).
